Very simplified likelihood

### 2024-03-19 inst/hankin-ejs.tex
minor rewording and improvement to figures

\begin{abstract}
Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Placket-Luce strength of $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from Formula~1 motor racing
and community initiative Parkrun.
\end{abstract}

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\cite{zipf1949} [a typical assumption for Bradley-Terry
  strengths], Hankin~\cite{hankin2017_rmd} shows that maximum
likelihood strengths identify the correct ranking with a probability
of essentially zero.  The situation is worse in cases such as the
International Mathematical Olympiad, which requires us to consider
over 100 competitors.

@@

Using $\mathcal{S}=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
$\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have
$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
alternatively $\hat{a}=\frac{n}{n+r+1}$.  For repeated observations
$r_i,n_i$, $1\leqslant i\leqslant m$ we would have
\begin{equation}
  \hat{a} = 1-
  \frac{\frac{r_1+1}{n_1}+\cdots+\frac{r_m+1}{n_m}}{m}
  = 1-\overline{\left(\frac{r+1}{n}\right)}.
\end{equation}

(that is, one minus the mean ratio of rank to number of [cloned]
competitors).  Previous related work presented the case $n=2$ [in
  current notation].  Considering the case $n=7$ gives us $7+1=8$
distinct likelihood functions, Figure~\ref{ninelikes}.  We see an
unanticipated asymmetry between $r$ and $n-r$.  For $r=4$ one might
expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
competitor's performance was median.  However, $\hat{a}\simeq 0.5886$.

@@

\section{Use-cases}

Here I present three examples that showcase likelihood
function~\ref{likeforrn3}.  Formula 1 education, parkrun

\subsection{Park run}

``Parkrun'' is a distributed community initiative that organises
weekly timed 5 km runs/walks in parks worldwide~\cite{hindley2020}.  A
typical event will have 200 participants.  Table~\ref{parkruntable}
shows the author has completed a total of 15 parkruns to date, and
from the first pair of numbers we see that 238 runners attended that
particular parkrun, of whom the author placed 173.

\begin{table*}[t]
  \caption{Parkrun results}
\label{parkruntable}
\begin{tabular}{ccccccccccc}
\hline
rank   &173  & 165 & 172 & 199 & 181 & 229 & 177 & 222 & 206 & 142 \\
runners&238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
rank   & 118 & 224 & 128 & 115 & 183 &     &     &     &     &\\
runners& 179 & 338 & 203 & 245 & 254 &     &     &     &     &\\
\hline
\end{tabular}
\end{table*}

We may consider the author to have an unknown generalized Bradley
Terry strength $a$: Figure~\ref{parkrunsupport} shows a support curve
for $a$ with the evaluate, $\hat{a}\simeq 0.448$ shown.  We also see
two units of support~\cite{edwards1992} shown as a dotted line
illustrating a support interval of about 0.315 - 0.567; we may be
reasonably confident that the author's true strength lies in this
range.  In particular, note that a reasonable $H_0\colon
a=\frac{1}{2}$ may not be rejected, the support for $H_0$ being only
susceptible to very minor improvement [$\simeq 0.351$] by moving to
the evaluate.

\begin{figure}[t]
\includegraphics[width=4in]{plotparkrun}  % plotparkrun.pdf made in very_simplified_likelihood.Rmd
\caption{Author's strength\label{parkrunsupport}}
\end{figure}

@@

One issue encountered in educational research is that of
confidentiality of examination scores, and individual results are
generally regarded as sensitive and private information.
Nevertheless, it is common for a student to possess some informative
observations: the total enrollment in a specific class and their
individual ranking within that class; table~\ref{educationtable} shows
an apocryphal dataset.  What could a student infer from such data?

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the
diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$.  Edwards's two units of
support criterion is met, and we are justified in concluding that the
student is indeed more competent at applied than pure mathematics.
Alternatively, we may observe that $2\mathcal{S}=4.86$ is in the tail
region of its asymptotic $\chi^2_1$ distribution, corresponding to a
$p$-value of about $0.028$.

@@

\begin{figure}[t]
\includegraphics[width=4in]{showchecolike}  % showchecolike.pdf produced by very_simplified_likelihood.Rmd
\caption{Likelihood functions for $\alpha,\beta$ for Checo's racing
  career 2011-2023\label{showchecolike}}
\end{figure}

>>>>

\begin{abstract}
Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Placket-Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from the fields of Olympic
athletics, education, and Formula~1 motor racing.
\end{abstract}

@@

\noindent where the strengths are normalised so $\sum p_i=1$.
However, estimation of the strengths is somewhat difficult if $n$ is
even moderately large: even with as few as 20 competitors following
Zipf's law~\cite{zipf1949} [a typical assumption for Bradley-Terry
  strengths], Hankin~\cite{hankin2017_rmd} shows that maximum
likelihood strengths identify the correct ranking with a probability
of essentially zero.  The situation is worse in cases such as formula
1 motor racing considered below, which requires us to consider
${\mathcal O}(100)$ competitors.

@@

Using $\mathcal{S}=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
$\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have
$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
alternatively $\hat{a}=\frac{n}{n+r+1}$.  Compare this with the result
of asking ``what is the probability $p$ that the focal competitor
beats a randomly chosen opponent?'' to which the binomial maximum
likelihood estimate is simply $\frac{n-r}{n}$, agreeing to within
$\mathcal{O}(n^{-2})$.  For repeated observations $r_i,n_i$,
$1\leqslant i\leqslant m$ we would have
\begin{equation}
  \hat{a} = 1-
  \frac{\frac{r_1+1}{n_1}+\cdots+\frac{r_m+1}{n_m}}{m}
  = 1-\overline{\left(\frac{r+1}{n}\right)}.
\end{equation}

(that is, one minus the mean ratio of rank to number of [cloned]
competitors); compare the binomial result of $1-\left.\sum
r_i\right/\sum n_i$.

Previous related work presented the case $n=2$ [in current notation].
Considering the case $n=7$ gives us $7+1=8$ distinct likelihood
functions, Figure~\ref{ninelikes}.  We see an unanticipated asymmetry
between $r$ and $n-r$.  For $r=4$ one might expect that
$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
performance was median.  However, $\hat{a}\simeq 0.5886$.

@@

Below, I present three examples that showcase likelihood
function~\ref{likeforrn3}, drawn from the fields of athletics,
education, and Formula~1 motor racing.

\subsection{Olympic track and field athletics}

One long-standing contention in the athletic world is that of lane
bias~\cite{munro2022}.  Running in the inner lane is frequently held
to be a disadvantage; mechanisms include biomechanical issues arising
from centrifugal acceleration, to competitors' inability to see their
rivals.  Here attention is restricted to heats, for which lane placing
is random.  I consider every heat in Olympic running from 2004-2020 (a
total of 41 heats, each of 8 or 9 competitors); only competitors'
ranks are used.  Figure~\ref{plotolymp} shows a log-likelihood curve
for the Bradley-Terry strength of the focal competitor, here the
runner in the innermost lane.  We see that the evaluate of $0.574$ is
very close to the neutral value of $0.5$, for which the support is
only $0.16$, far short of Edwards's two units of support criterion for
rejection~\cite{edwards1972}.  This dataset contains no evidence to
support lane bias.

\begin{figure}[t]
\includegraphics[width=4in]{plotolymp} % produced by very_simplified_likelihood.Rmd
\caption{Log-likelihood curve for the Bradley-Terry strength of the
  lane-1 runner, olympic qualifying heats 2012-2020 \label{plotolymp}}
\end{figure}

@@

In educational research, confidentiality of examination scores is
often an issue, and individual results are generally regarded as
sensitive and private information.  Nevertheless, it is common for a
student to possess some informative observations: the total enrollment
in a specific class and their individual ranking within that class;
table~\ref{educationtable} shows an apocryphal dataset.  What could a
student infer from such data?

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$.  Edwards's two units of
support criterion~\cite{edwards1972} is met, and we are justified in
concluding that the student is indeed more competent at applied than
pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

@@

\begin{figure}[t]
\includegraphics[width=4in]{showchecolike}  % showchecolike.pdf produced by very_simplified_likelihood.Rmd
\caption{Log-likelihood contours for $\alpha,\beta$ for Checo's racing
  career 2011-2023\label{showchecolike}}
\end{figure}

### 2024-03-19 inst/very_simplified_likelihood.Rmd
minor rewording and improvement to figures

```{r showfilledc}
L <- pmax(L,-40)
showchec <- function(...){
contour(a,b,L,xlab=expression(alpha),ylab=expression(beta))
}
showchec()
pdf(file="showchecolike.pdf")
contour(a,b,L)
dev.off()
```

@@

```{r plotolymp}
plot(a,L,type='l',ylab="log-likelihood",lwd=2)
abline(h=c(0,-2))
ahat <- optolym$maximum
segments(x0=ahat,y0=-1.5,y1=0)
text(ahat,-0.91,expression(hat(a)==0.573),pos=2)
abline(v=0.5,lty=3)

@@

text(x=a_lower,y=-3,"0.380",pos=4)
text(x=a_upper,y=-3,"0.563",pos=2)
```

>>>>

```{r showfilledcwithdot}
L <- pmax(L,-40)
showchec <- function(...){
contour(a,b,L,xlab=expression(alpha),ylab=expression(beta),levels=-c(2,5*(1:5)))
points(-0.27,0.0813,pch=16,cex=2)
}
showchec()
pdf(file="showchecolike.pdf")
showchec(a,b,L)
dev.off()
```

@@

```{r plotolymp}
plotolymp <- function(...){
plot(a,L,type='l',ylab="log-likelihood",lwd=2)
abline(h=c(0,-2))
ahat <- optolym$maximum
segments(x0=ahat,y0=-1.5,y1=0)
text(ahat,-0.91,expression(hat(a)==0.573),pos=2)
abline(v=0.5,lty=3)

@@

text(x=a_lower,y=-3,"0.380",pos=4)
text(x=a_upper,y=-3,"0.563",pos=2)
}
plotolymp()
pdf(file="plotolymp.pdf")
plotolymp()
dev.off()
```
